Measurement Bias from Address Aliasing

### 2014-05-26 paper.tex
Merge malloc/allocator sections

\subsection{Why \texttt{malloc} will Alias by Default}
Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
Depending on the particular request and allocator used, the returned value will either point to the ``regular'' heap, or to a memory mapped area (see Figure \ref{fig:virtualmemory}).
Different strategies for how to allocate memory will often depend on the \emph{size} of the request.

\begin{enumerate}
  \item The heap is marked by a brake point representing the end of uninitialized data in virtual memory, and more space is requested by the \texttt{brk} or \texttt{sbrk} system calls. 
  Allocators typically use this area for smaller allocations.
  \item Anonymous memory mappings, i.e. buffers not backed by a file, can be created through the \texttt{mmap} system call.
  This mechanism is typically used for large allocations.
  %For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
\end{enumerate}
% One drawback of using memory mapping is that it requires the buffer to be zero initialized, so which mechanism is used can be based on some cost heuristic employed by the allocator.
% A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
% The normal heap is often used for smaller requests, while memory mapping is used for large buffers.

The heap section starts at a relatively low address right above static code and data.
Memory mapped chunks are placed towards the upper end of the virtual address space, closer to stack.
Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the actual pointer values returned.
While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.

\subsection{Aligned Sequential Access}
Many functions operate in a ``sliding window'' fashion; reading and writing to different buffers in some loop construction.
This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
As an example of this, consider a naive implementation of linear correlation shown in Figure \ref{lst:conv}.
The function containins lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
We will see that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.

@@

By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
To analyse performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the arguments before calling the function.
Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
Allocating and managing these buffers at startup takes a non-neglegible amount of work.
The overhead can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
\begin{lstlisting}
    for (i = 0; i < k; ++i)
        convolution(n, input, output + offset);
\end{lstlisting}
An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.

\[
t_{\text{estimate}} = \frac{t_{k} - t_{1}}{k - 1}
\]

Our results are with $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
For every run, performance counter measurements is also averaged over 10 samples, using the repeat mechanism of perf.

@@

\subsection{Behavior of other Allocators}
As with the contents of environment variables, the exact behavior of dynamic memory allocation can typically not be known ahead of time.
Heap allocation routines such as \texttt{malloc} and \texttt{free} are usually dynamically linked and resolved at runtime, which again represents a potential source of bias.
For the code examples presented earlier, the dynamic linker will resolve \texttt{malloc} to some version of glibc, which is the default on our system.
Variations between different versions of libc, or local configurations on parameters such as thresholds for when to use \texttt{mmap} over \texttt{brk}, could trigger aliasing cases.

Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
\begin{itemize}
  \item Thread-Caching Malloc (tcmalloc) from Google. It claims to be time- and space efficient, and reduces lock contention for multithreaded applications~\cite{TCMalloc}.
  \item jemalloc was originally developed for FreeBSD. It aims to be a scalable general purpose allocator in single or multithreaded environments~\cite{JEMalloc}.
  \item Hoard is another project that focuses extensively on improving performance for multithreaded programs~\cite{Berger:2000:Hoard}.
\end{itemize}
The addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} are observed for different size parameters. 
We do not care about performance, only whether the actual memory addresses returned alias or not.

\begin{table}[t]
  \label{tab:mallocompare}
  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
    \pgfplotstabletypeset[
      col sep=comma,
      string type,
      columns={Allocation, 64, 5120, 1048576},
      column type=r,
      columns/Allocation/.style={
        string type, 
        column type=l,
        column type/.add={|}{}
      },
      every head row/.style={
        output empty row,
        before row={\hline
           & 64 B & 5,120 B & 1,048,576 B \\
          },
        after row=\hline\hline,
      },
      every last row/.style={after row=\hline},
      every last column/.style={column type/.add={}{|}},
      every odd row/.style={after row=\hline},
    ]{bin/malloc-comparison.csv}
  }
  \begin{tabnote}
    \Note{Note:}{Memory mapped addresses starting with the 0x2aa... prefix is an artifact of using \emph{make} to generate results. Executing the test program from \emph{bash} directly result in 0x7fff... prefixes.
    The difference is not important to our discussion.}
  \end{tabnote}
\end{table}

For smaller allocations, we see that glibc and tcmalloc uses the normal heap area -- returning numerically low addresses.
Interestingly, jemalloc and Hoard appears to never use the normal heap, but allocate to memory mapped areas even for smaller sizes.
We see one example where one allocator yields aliasing buffers while another does not.
Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.

%The purpose of this is to illustrate that allocators is yet another source of bias.
Using heap allocated memory, address aliasing is an artifact of the allocator used.
With different versions or configurations of memory allocators, any performance impact from aliasing can qualify as measurement bias.

% Most malloc alternatives focuses heavily on performance in multithreaded environments.
% Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.

>>>>

\subsection{Most Allocators Alias by Default}
Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
Depending on the particular request and allocator used, the returned value will either point to the ``regular'' heap, or to a memory mapped area (see Figure \ref{fig:virtualmemory}).
Different strategies for how to allocate memory will often depend on the \emph{size} of the request.

\begin{enumerate}
  \item The heap is marked by a brake point representing the end of uninitialized data in virtual memory, and more space is requested by the \texttt{brk} or \texttt{sbrk} system calls. 
  Allocators typically use this area for smaller allocations.
  \item Anonymous memory mappings, i.e. buffers not backed by a file, can be created through the \texttt{mmap} system call.
  This mechanism is typically used for large allocations.
\end{enumerate}

The heap section starts at a relatively low address right above static code and data.
Memory mapped chunks are placed towards the upper end of the virtual address space, closer to stack.
Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the pointer values returned:
While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger, for example 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{glibc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address ends with 0x010.}
This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.

Table \ref{tab:mallocompare} illustrates how different dynamic allocators tend to produce aliasing conditions for large requests\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
The addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} are observed for different size parameters, and equal three digit address suffix indicate an aliasing pair.
In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
\begin{itemize}
  \item Thread-Caching Malloc (tcmalloc) by Google~\cite{TCMalloc}.
  \item jemalloc, originally developed for FreeBSD~\cite{JEMalloc}.
  \item Hoard~\cite{Berger:2000:Hoard}.
\end{itemize}
All of the above focuses heavily on performance in multithreaded environments.
Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
Here we only look at behavior for a single thread, and whether the addresses returned alias or not.
\begin{table}[t]
  \label{tab:mallocompare}
  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
    \pgfplotstabletypeset[
      col sep=comma,
      string type,
      columns={Allocation, 64, 5120, 1048576},
      column type=r,
      columns/Allocation/.style={
        string type, 
        column type=l,
        column type/.add={|}{}
      },
      every head row/.style={
        output empty row,
        before row={\hline
           & 64 B & 5,120 B & 1,048,576 B \\
          },
        after row=\hline\hline,
      },
      every last row/.style={after row=\hline},
      every last column/.style={column type/.add={}{|}},
      every odd row/.style={after row=\hline},
    ]{bin/malloc-comparison.csv}
  }
  \begin{tabnote}
    \Note{Note:}{Memory mapped addresses starting with the 0x2aa... prefix is an artifact of using \emph{make} to generate results. Executing the test program from \emph{bash} directly result in 0x7fff... prefixes.
    The difference is not important to our discussion.}
  \end{tabnote}
\end{table}

For smaller allocations, we see that glibc and tcmalloc utilize the normal heap area -- returning numerically low addresses.
Interestingly, jemalloc and Hoard appears to never use the heap, but allocate to memory mapped areas even for smaller requests.
Conversely, tcmalloc seem manage only the heap.
We see one example where one allocator yields aliasing buffers while another does not.
Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.
But even with randomization, addresses returned by \texttt{mmap} must still be page aligned.
This means that addresses returned by allocators using this mechanism directly will \emph{always} alias, giving a deterministic execution context considering only the address suffix.
From our limited experiment, this seems to be the case for glibc, jemalloc and Hoard.

\subsection{Aligned Sequential Access}
Many functions operate in a ``sliding window'' fashion; reading and writing to different buffers in some loop construction.
This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
As an example of this, consider a naive implementation of linear correlation shown in Figure \ref{lst:conv}.
The function contains lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
We will see that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.

@@

By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
To analyse performance for different addresses, we manually insert padding to offset one of the memory mapped pointers.
This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset one of the function arguments.
Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
Allocating and managing these buffers at startup takes a non-neglegible amount of work.
The overhead can be masked by repeatedly invoking the convolution kernel after allocating and initializing all the inputs.
\begin{lstlisting}
    for (i = 0; i < k; ++i)
        conv(n, input, output + offset);
\end{lstlisting}
An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.
\[
t_{\text{estimate}} = \frac{t_{k} - t_{1}}{k - 1}
\]
Our results are with $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
For every run, performance counter measurements is also averaged over 10 samples, using the repeat mechanism of perf.

@@
